Towards a Verified Bzip2: A Functional Specification for Burrows--Wheeler

### 2019-05-06 draft_paper/draft_paper.tex
stash

The forwards transform is straightforward to describe: for
a string of length \(n\), form the \(n \times n\) rotaion matrix of all
cyclic shifts of the original string, sort it lexicographically, and
take the last column of the matrix. In addition, we note the index
where the original string appears in the sorted rotation matrix. As
shown in \Cref{fig:bw_ex}, applying the algorithm to the string
\var{abracadabra!} produces the new string \var{ard!rcaaabb} and
the index \var{3}. The transformation tends to produce runs like
the three \var{a}s or the two \var{b}s in \var{ard!rcaaabb},
and it is this feature that makes the resulting string easier to
compress. Although it is not obvious, the transformed string, together
with the index, provide enough information to recover the the original
string. The BWT is discussed more in \cref{sec:bwt}.

@@

Our main contribution is a formalization in the Coq proof assistant of
the BWT, together with a machine-checked proof that our implementation
is correct. Specifically, we implement the functions $\var{bwp} :
\var{list}~A \to \var{list}~A$ and $\var{bwn} : \var{list}~A
\to \mathbb{N}$ that produce the string and index part of the BWT
respectively, and the function $\var{unbwt} : \mathbb{N} \to
\var{list}~A \to \var{list}~A$ that reverses the transform. We
prove the following theorem:
\begin{theorem}[unbwt\_correct]
  \label{thm:unbwt_correct}
  For all lists $l$,
  \begin{equation*}
    \var{unbwt}~(\var{bwn}~l)~(\var{bwp}~l) = l
  \end{equation*}
\end{theorem}
The implementation of the functions and the outline of the proof are
based on Bird et~al.'s derivations.

@@

As a part of the proof of \Cref{thm:unbwt_correct}, we implemented
radix-sort and proved its correctness,
\begin{theorem}[radixsort\_correct]
  \label{thm:radixsort_correct}%
  For all lists $l$,
  \begin{equation*}
    \var{Sorted}~(\var{radixsort}~l) \land \var{Permutation}~(\var{radixsort}~l)~l
  \end{equation*}
\end{theorem}
and proved that any two stable sorts are interchangable,
\begin{theorem}[stable\_sort\_unique]
  \label{thm:radixsort_correct}%
  For all functions $f, g : \var{list}~A \to \var{list}~A$,
  \begin{gather*}
    (\forall l,~\var{Sorted}~(f~l) \land \var{StablePermutation}~(f~l)~l) \implies \\
    (\forall l,~\var{Sorted}~(g~l) \land \var{StablePermutation}~(g~l)~l) \implies \\
    \forall l,~f~l = g~l
  \end{gather*}
\end{theorem}
This in turn required developing suitable definitions for
\var{Sorted}, \var{Permutation}, and \var{StablePermutation}.
We explored multiple definitions for each of these properties, and
proved equivalence between them.

The rest of this paper is organize as follows. \Cref{sec:bwt} details
the history of the BWT and its applications. \Cref{sec:ord} defines
what we need to know about a type $A$ to sort a $\var{list}~A$ as a
part of BWT. Following Bird et~al., \cref{sec:forwards_BWT} defines
the implementation of the forwards BWT in Coq, and
\cref{sec:inverse_BWT} derives an implementation of the inverse BWT
via program calculation. \Cref{sec:sorted_perm_stable} provides
multiple definitions of \var{Sorted}, \var{Permutation}, and
\var{StablePermutation}, and proves equivalence between them.
\Cref{sec:sort} implements insertion sort and radixsort, and proves
their correctness. \Cref{appendix:BWT_proof} more formally proves the
theorems about the BWT mentioned in previous sections, including
\Cref{thm:unbwt_correct}.

@@

\section{Burrows--Wheeler}
\label{sec:bwt}

@@

The next pass is a Move-to-Front (MTF) encoding, in which a character
is represented by the number of different characters encountered since
the last occurence of this character. This creates many runs of zeroes
from the runs in the output of BWT. On English texts, often 50-60\% of
the values produced by the MTF pass are zeroes
\cite{fenwick2007,bw-analysis}. Often a Run-Length Encoding (RLE) pass
is then used to encode only the runs of zeroes with their lengths,
using a bijective base-2 encoding \cite{bw-analysis, tsai_2016}.
Finally, some sort of zeroth-order coder like a Huffman coder is used.

BWT is also at the heart of algorithms that exploit the relationship
between the rotation matrix in BWT and the suffix array data structure
in order to search for patterns in a compressed text
\cite{ferragina_index}.

@@

Although it is commonly demonstrated on strings, the BWT can be
applied to any list whose elements are comparable. To make that
precise, if we have a type $A$ and a relation on that type $\le$, we say
that this forms a preorder on $A$ when $\le$ is reflexive and transitive, that is
$x \le x$ for all $xA$, and
\begin{equation*}
  x \le y \implies y \le z \implies x \le z
\end{equation*}
for all $x,y,z$.

@@

From $\le$, we can define $x \ge y \coloneqq y \le x$, $x < y \coloneqq
\lnot (y \le x)$, and $x > y \coloneqq \lnot (x \le y)$, and prove
transitivity and reflexivity where appropriate.

We can also define $x \equiv y \coloneqq x \le y \land y \le x$. $\equiv$ is a
decidable equivalence relation, which means it is reflexive,
transitive, and symmetric, and we have a decision procedure for it.
\lstinputlisting[
  firstline=116, lastline=120, widthgobble=2
]{../theories/BWT/Sorting/Ord.v}

The builtin class \var{EqDec} is used to describe types with
decidable equivalence relations, so for any \var{Preord} instance
we have the \var{EqDec} instance \var{Preord\_EqDec} using
\var{eqv\_dec}. In Coq, the notation \var{===} is used for the
equivalence relation of an \var{EqDec} instance and \var{==} for
its decision procedure.

@@

We can prove that this definition satisfies \Cref{eq:iter} by defining
the function \var{rep} for repeated function application.
\lstinputlisting[
  widthgobble=2,
  firstline=8,
  lastline=12
]{../theories/BWT/Lib/Repeat.v}
In proofs, we will still use the notation $f^i z$ for $\var{rep}~f~i~z$.

@@

\var{bwp} produces the transformed string by taking the last column
of the sorted rotation matrix. \var{lexsort} sorts the rotation
matrix lexicographically using the \var{Ord} instance; we will
discuss it's implementation in \cref{sec:sort}.
\lstinputlisting[
  widthgobble=2,
  firstline=31,
  lastline=35
]{../theories/BWT/BurrowsWheeler.v}

@@

Now we want to construct a function \var{unbwt} such that for all $l$,
\begin{equation}
  \var{unbwt}~(\var{bwn}~l)~(\var{bwp}~l) = l
  \label{eq:unbwt}
\end{equation}
If we could recreate the entire sorted rotation matrix from
$\var{bwp}~l$, it would be simple to recover $l$ by indexing into
the matrix at $\var{bwn}~l$.

More precisely, suppose we had a function function \var{recreate}
such that
\begin{equation}
  \var{recreate}~(\var{bwp}~l) = \var{lexsort}~(\var{rots}~l)
  \label{eq:recreate}
\end{equation}
Then, we could define \var{unbwt} as
\begin{lstlisting}
Definition unbwt (i : nat) (l : list A) : list A :=
  nth i (recreate l) l.
\end{lstlisting}
and \Cref{eq:unbwt} would follow directly from the \var{bwn\_correct}:
\begin{align*}
     \var{unbwt}~(\var{bwn}~l)~(\var{bwp}~l)
  &= (\var{recreate}~(\var{bwp}~l))_{(\var{bwn})~l}\\
  &= (\var{lexsort}~(\var{rots}~l))_{(\var{bwn})~l} \\
  &= l
\end{align*}

@@

\begin{theorem}[recreate\_correct\_inv]
  For all lists $l$ and $j \le \var{length}~l$,
  \begin{equation*}
    \var{recreate}~j~(\var{bwp}~l) =
    \var{cols}~j~(\var{lexsort}~(\var{rots}~l))
  \end{equation*}
\end{theorem}

@@

Now we are ready to prove the invariant:
\begin{theorem}[recreate\_correct\_inv]
  For all lists $l$ and $j \le \var{length}~l$,
  \begin{equation*}
    \var{recreate}~j~(\var{bwp}~l) =
    \var{cols}~j~(\var{lexsort}~(\var{rots}~l))
  \end{equation*}
\end{theorem}
\begin{proof}
\begin{align*}
  &\var{cols}~(j+1)~(\var{lexsort}~(\var{rots}~l)) \\
={}& \var{cols}~(j+1)~(\var{hdsort}~(\var{map}~\var{rrot}~(\var{lexsort}~(\var{rots}~l)))) \\
={}&
\var{hdsort}~(\var{cols}~(j+1)~(\var{map}~\var{rrot}~(\var{lexsort}~(\var{rots}~l)))) \\
={}&
\var{hdsort}~(\var{prepend\_col}~(\var{map}~\var{last}~(\var{lexsort}~(\var{rots}~l)))~(\var{cols}~j~(\var{lexsort}~(\var{rots}~l))) \\
={}&
\var{hdsort}~(\var{prepend\_col}~(\var{bwp}~l)~(\var{recreate}~j~(\var{bwp}~l))) \\
={}& \var{recreate}~j~(\var{bwp}~l) \qedhere
\end{align*}
\end{proof}

@@

The previous section's derivation relied on the correctness and stability of
various sorts. To make those parts of the derivation precise, we
define here what it means for a sorting function to be correct, and
what it means for a sort to be stable.

Intuitively, a sorting function must produce a sorted list that
contains all of the elements that its input contained. So, given a function $f : \var{list}~A \to \var{list}~A$, we say that $\var{Sort}~f$ precisely when for all $l$,
\begin{equation*}
  \var{Sorted}~(f~l) \land \var{Permutation}~(f~l)~l
\end{equation*}
Now we need to define \var{Sorted} and \var{Permutation} to capture
the idea that a list is in sorted order, and that two lists contain
all of the same elements.

@@

However, it is often easier to reason about inductively defined
properties, so we would like to define \var{Sorted} inductively. So,
we can say that as a base case, the empty list is sorted, and given a
sorted list, we can add an element that is smaller than all the
existing elements and the result will also be sorted:
\begin{align*}
  \infer[\var{Sorted\_nil}]{\var{Sorted}~[]}{}
  &&
  \infer[\var{Sorted\_cons}]{\var{Sorted}~(a :: l)}{%
    (\forall~x \in l,~a < x) & \var{Sorted}~l
  }
\end{align*}

@@

Although these two inductive definitions are intuitive, they are not
obviously equivalent to the first definition, which is more clearly
correct. We prove them equivalent all:
\begin{theorem}[SortedLocal\_iff]
  For all lists $l$,
  \begin{equation*}
    \var{Sorted}~l \iff \var{SortedLocal}~l
  \end{equation*}
\end{theorem}
\begin{proof}
  The $\implies$ direction follows by induction. The $\impliedby$
  direction follows by induction on $l$, reasoning about the empty
  list, the singleton list, and lists with at least two elements. In
  the inductive case $l = a :: b :: l'$, we have that $a \le b$, and
  that $\forall x \in l',~b \le x$, which gives us $\forall x \in b :: l',~a \le x$ by
  transitivity.
\end{proof}

\begin{theorem}[SortedIx\_iff]
  A list $l$ is sorted under the inductive definition (\var{Sorted}) if
  and only if it is sorted under the index definition (\var{SortedIx}).
\end{theorem}
\begin{proof}
  We reason by induction on $l$. The base cases are easy, so we
  address the inductive case, $l = a :: l'$

  ($\implies$) We have to prove that $(a :: l')_i \le (a :: l')_j$ given
  $i \le j$ and $\var{Sorted}~l$. We reason by cases on $i$ and $j$:
  \begin{itemize}[noitemsep]
  \item $i = j = 0$. By reflexivity, $a \le a$.
  \item $i = 0,~j=j' + 1$. We need to show $a \le l'_{j'}$. We know that
    $a \le x$ for all $x \in l'$ by $\var{SortedIx}~(a::l')$, so we are
    done.
  \item $i = i' +1,~j = j' + 1$. We need to show $l'_{i'} \le l'_{j'}$,
    which follows from the induction hypothesis.
  \end{itemize}

@@

The word \textit{permutation} has two related meanings that are often
used interchangably. We sometimes say that two lists are permutations of each
other if they contain all of the same elements, just appearing in a
different order. However, in mathematics we often say that a
permutation is actually the bijection on indices that defines the
correspondence between the two lists. For example, we say that $p$ is
the permutation that relates $l$ to $l'$ if for all indices $i$ and
$j$,
\begin{equation*}
  l_{p(i)} = l'_i
\end{equation*}
For consistency with Coq's built-in \var{Permuation} relation, in a
situation like this we will say that $l$ and $l'$ are permutations of
each other, and that $p$ is the \textit{permutation function} that
relates $l$ to $l'$. We say that two lists $l$ and $l'$ are
permutations of each other is precisely when
\vspace{-\parskip}
\begin{itemize}[nolistsep]
  \item $\var{length}~l = \var{length}~l' = n$
  \item there exists a permutation function $p : \mathbb{N}_{<n} \to
    \mathbb{N}_{<n}$, where $p$ is a bijection, such that for all indices $i$,
    \begin{equation*}
      l_{p(i)} = l'_i
    \end{equation*}
\end{itemize}

@@

We will prove these two definitions equivalent, but first we confirm that our definition of \var{PermFun} ensures that .

@@

\begin{corollary}[StableSort\_unique]
  For all functions $f, g: \var{list}~A \to \var{list}~A$,
  \begin{equation*}
    \var{StableSort}~f \implies \var{StableSort}~g \implies
    (\forall l,~f~l = g~l)
  \end{equation*}
\end{corollary}
\begin{proof}
  This follows from \var{StablePerm\_Sorted\_eq}.
\end{proof}

@@

\begin{corollary}[Sort\_Ord\_unique]
  When sorting using an order,
  \begin{equation*}
    \var{Sort}~f \implies \var{Sort}~g \implies
    (\forall l,~f~l = g~l)
  \end{equation*}
  for any $f$ and $g$.
\end{corollary}
\begin{proof}
  This follows from \var{Sort\_StableSort\_Ord} and \var{StableSort\_unique}.
\end{proof}

@@

%% \begin{proof}
%%   We reason by induction on $m$. In the inductive case, we need to show
%%   \begin{equation*}
%%     \var{PrefixSorted}~(j+1)~(\var{hdsort}~(r::m))
%%   \end{equation*}
%%   If we simplify \var{hdsort}, this becomes
%%   \begin{equation*}
%%     \var{PrefixSorted}~(j+1)~(\var{insert}_1~r~(\var{hdsort}~m))
%%   \end{equation*}
%%   where $\var{insert}_1$ is \var{insert} using the $1$-prefix
%%   lexicographic ordering used in \var{hdsort}.
%%   We know from our assumption that
%%   \begin{equation*}
%%     \var{PrefixSorted}~j~(\var{tl}~x :: \var{map}~\var{tl}~m)
%%   \end{equation*}
%%   and from our induction hypothesis we have
%%   \begin{equation*}
%%     \var{PrefixSorted}~(j+1)~(\var{hdsort}~m)
%%   \end{equation*}
%% \end{proof}

  %% We can deconstruct $\var{insert}_1~r~(\var{hdsort}~m)$ as
  %% $m_1 \dplus r :: m_2$, where we also know that $m = m_1 \dplus m_2$.
  %% Now we have to show
  %% \begin{equation*}
  %%   \var{PrefixSorted}~(j+1)~(m_1 \dplus r :: m_2)
  %% \end{equation*}

>>>>

The forwards transform is straightforward to describe: for a string of
length \(n\), form the \(n \times n\) matrix of all cyclic shifts of
the original string (also referred to as the rotation matrix), sort
the rows lexicographically, and take the last column of the matrix. In
addition, we note the index where the original string appears in the
sorted rotation matrix. As shown in \Cref{fig:bw_ex}, applying the
algorithm to the string \var{abracadabra!} produces the new string
\var{ard!rcaaabb} and the index \var{3}. The transformation tends to
produce runs like the three \var{a}s or the two \var{b}s in
\var{ard!rcaaabb}, and it is this feature that makes the resulting
string easier to compress. Although it is not obvious, the transformed
string, together with the index, provide enough information to recover
the the original string. The BWT is discussed more in \cref{sec:bwt}.

@@

Our main contribution is a formalization in the Coq proof assistant of
the BWT, together with a machine-checked proof that our implementation
is correct. Specifically, we implement the functions $\var{bwl} :
\var{list}~A \to \var{list}~A$ and $\var{bwn} : \var{list}~A \to
\mathbb{N}$ that produce the string and index part of the BWT
respectively, and the function $\var{unbwt} : \mathbb{N} \to
\var{list}~A \to \var{list}~A$ that reverses the transform. We prove the
following theorem:
\begin{theorem}[unbwt\_correct]
  \label{thm:unbwt_correct}
  For all lists $l$,
  \begin{equation*}
    \var{unbwt}~(\var{bwn}~l)~(\var{bwl}~l) = l
  \end{equation*}
\end{theorem}
The implementation of the functions and the outline of the proof are
based on Bird et~al.'s derivations.

@@

As a part of the proof of \Cref{thm:unbwt_correct}, we proved two
other noteworthy results that have not, to our knowledge, been proved
before in Coq. We implemented a least-significant-digit (LSD)
radixsort and proved its correctness,
\begin{theorem}[radixsort\_correct]
  \label{thm:radixsort_correct}%
  For all matrices $m$ with $n$ columns,
  \begin{equation*}
    \var{Sorted}~(\var{radixsort}~m~n) \land \var{Permutation}~(\var{radixsort}~m~n)~m
  \end{equation*}
\end{theorem}
and we proved that any two stable sorts are interchangable,
\begin{restatable*}[StableSort\_unique]{theorem}{stablesortunique}
  For all functions $f, g: \var{list}~A \to \var{list}~A$,
  \begin{equation*}
    \var{StableSort}~f \implies \var{StableSort}~g \implies
    (\forall l,~f~l = g~l)
  \end{equation*}
\end{restatable*}
This in turn required developing suitable definitions for
\var{Sorted}, \var{Permutation}, and \var{StablePerm}. We explored
multiple definitions for each of these properties, and proved
equivalence between them.

The rest of this paper is organized as follows. \Cref{sec:bwt} details
the history of the BWT and its applications. \Cref{sec:ord} defines
what we need to know about a type $A$ to sort a $\var{list}~A$ as a
part of BWT. Following Bird et~al., \cref{sec:forwards_BWT} defines
the implementation of the forwards BWT in Coq, and
\cref{sec:inverse_BWT} derives an implementation of the inverse BWT
via program calculation. \Cref{sec:sorted_perm_stable} provides
multiple definitions of \var{Sorted}, \var{Permutation}, and
\var{StablePermutation}, and proves equivalence between them.
\Cref{sec:sort} implements insertion sort and radixsort, and proves
their correctness.

@@

\section{History and Applications of the BWT}
\label{sec:bwt}

@@

The next pass is a Move-to-Front (MTF) encoding, in which a character
is represented by the number of different characters encountered since
the last occurence of this character. This creates many runs of zeroes
from the runs in the output of BWT. On English texts, often 50-60\% of
the values produced by the MTF pass are zeroes \cite{fenwick2007}.
Often a Run-Length Encoding (RLE) pass is then used to encode only the
runs of zeroes with their lengths, using a bijective base-2 encoding
\cite{bw-analysis, tsai_2016}. Finally, some sort of zeroth-order
coder like a Huffman coder is used.

BWT is also at the heart of algorithms that exploit the relationship
between the rotation matrix in BWT and the suffix array data structure
in order to search for patterns in a compressed text
\cite{ferragina_index}. These techniques have been applied to enable
fast matching of DNA subsequences.

@@

Although it is commonly demonstrated on strings, the BWT can be
applied to any list whose elements are comparable. To make that
precise, if we have a type $A$ and a relation on that type $\le$, we say
that this forms a preorder on $A$ when $\le$ is reflexive and transitive, that is
$x \le x$ for all $x:A$, and
\begin{equation*}
  x \le y \implies y \le z \implies x \le z
\end{equation*}
for all $x,y,z:A$.

@@

From $\le$, we can define
\begin{align*}
x \ge y \coloneqq y \le x && x < y \coloneqq \lnot (y \le x) && x > y \coloneqq \lnot (x \le y)
\end{align*}

We can also define
\begin{equation*}
  x \equiv y \coloneqq x \le y \land y \le x
\end{equation*}
$\equiv$ is a
decidable equivalence relation, which means it is reflexive,
transitive, and symmetric, and we have a decision procedure for it.

The builtin class \var{EqDec} is used to describe types with decidable
equivalence relations, so for any \var{Preord} instance we have the
\var{EqDec} instance \var{Preord\_EqDec}. In Coq, the notation
\var{===} is used for the equivalence relation of an \var{EqDec}
instance and \var{==} for its decision procedure.

@@

We can prove that this definition satisfies \Cref{eq:iter} by defining
the function \var{rep}\footnote{Alternatively, we can say that
  $\lambda f.~\var{rep}~n~f$ is the $n$th Church numeral.} for
repeated function application.
\lstinputlisting[
  widthgobble=2,
  firstline=8,
  lastline=12
]{../theories/BWT/Lib/Repeat.v}
In proofs, we will still use the notation $f^i z$ for $\var{rep}~f~i~z$.

@@

\var{bwl} produces the transformed string by taking the last column
of the sorted rotation matrix. \var{lexsort} sorts the rotation
matrix lexicographically using the \var{Ord} instance; we will
discuss it's implementation in \cref{sec:sort}.
\lstinputlisting[
  widthgobble=2,
  firstline=31,
  lastline=35
]{../theories/BWT/BurrowsWheeler.v}

@@

Now we want to construct a function \var{unbwt} such that for all $l$,
\begin{equation}
  \var{unbwt}~(\var{bwn}~l)~(\var{bwl}~l) = l
  \label{eq:unbwt}
\end{equation}
If we could recreate the entire sorted rotation matrix from
$\var{bwl}~l$, it would be simple to recover $l$ by indexing into
the matrix at $\var{bwn}~l$.

More precisely, suppose we had a function function \var{recreate}
such that
\begin{equation}
  \var{recreate}~(\var{bwl}~l) = \var{lexsort}~(\var{rots}~l)
  \label{eq:recreate}
\end{equation}
Then, we could define \var{unbwt} as
\begin{lstlisting}
Definition unbwt (i : nat) (l : list A) : list A :=
  nth i (recreate l) l.
\end{lstlisting}
and \Cref{eq:unbwt} would follow directly from the \var{bwn\_correct}:
\begin{align*}
     \var{unbwt}~(\var{bwn}~l)~(\var{bwl}~l)
  &= (\var{recreate}~(\var{bwl}~l))_{(\var{bwn})~l}\\
  &= (\var{lexsort}~(\var{rots}~l))_{(\var{bwn})~l} \\
  &= l
\end{align*}

@@

\begin{theorem}[recreate\_correct\_inv]
  For all lists $l$ and $j \le \var{length}~l$,
  \begin{equation*}
    \var{recreate}~j~(\var{bwl}~l) =
    \var{cols}~j~(\var{lexsort}~(\var{rots}~l))
  \end{equation*}
\end{theorem}

@@

Now we are ready to prove the invariant:
\begin{theorem}[recreate\_correct\_inv]
  For all lists $l$ and $j \le \var{length}~l$,
  \begin{equation*}
    \var{recreate}~j~(\var{bwl}~l) =
    \var{cols}~j~(\var{lexsort}~(\var{rots}~l))
  \end{equation*}
\end{theorem}
\begin{proof}
\begin{align*}
  &\var{cols}~(j+1)~(\var{lexsort}~(\var{rots}~l)) \\
={}& \var{cols}~(j+1)~(\var{hdsort}~(\var{map}~\var{rrot}~(\var{lexsort}~(\var{rots}~l)))) \\
={}&
\var{hdsort}~(\var{cols}~(j+1)~(\var{map}~\var{rrot}~(\var{lexsort}~(\var{rots}~l)))) \\
={}&
\var{hdsort}~(\var{prepend\_col}~(\var{map}~\var{last}~(\var{lexsort}~(\var{rots}~l)))~(\var{cols}~j~(\var{lexsort}~(\var{rots}~l))) \\
={}&
\var{hdsort}~(\var{prepend\_col}~(\var{bwl}~l)~(\var{recreate}~j~(\var{bwl}~l))) \\
={}& \var{recreate}~j~(\var{bwl}~l) \qedhere
\end{align*}
\end{proof}

@@

The previous section's proofs relied on the correctness and stability
of various sorts. The Coq standard library includes definitions for
\var{Sorted} and \var{Permutation}, but no definition for stability.
The Coq standard library also includes an implementation of mergesort.
However, it uses Coq's module system to enable parametricity over
ordered types, which is much less convenient to use than typeclasses.
In this section, we define our own version of \var{Sorted} and
\var{StablePerm} that are based on our \var{Preord} typeclass.

When defining these properties, it is very important that they mean
what we would like them to mean. If we prove our implementations
satisfy our specifications, but our specifications are incorrect, then
the proof is worthless. In addition, it would be nice if the
properties are defined in a way that is elegant to reason about. So,
we define each property in multiple ways and prove equivalence between
them: some definitions are more obviously correct, while others might
be easier to reason about it different situations. Even if a
definition is never used, proving equivalence gives us confidence
about the correctness of the definitions.

@@

However, it is often easier to reason about inductively defined
properties, so we would like to define \var{Sorted} inductively. We
can say that as a base case, the empty list is sorted, and given a
sorted list, we can add an element that is smaller than all the
existing elements and the result will also be sorted:
\begin{align*}
  \infer[\var{Sorted\_nil}]{\var{Sorted}~[]}{}
  &&
  \infer[\var{Sorted\_cons}]{\var{Sorted}~(a :: l)}{%
    (\forall~x \in l,~a < x) & \var{Sorted}~l
  }
\end{align*}

@@

Although these two inductive definitions are intuitive, they are not
obviously equivalent to the first definition, which is more clearly
correct. We will prove all three equivalent.

\begin{theorem}[SortedLocal\_iff]
  For all lists $l$,
  \begin{equation*}
    \var{Sorted}~l \iff \var{SortedLocal}~l
  \end{equation*}
\end{theorem}
\begin{proof}
  Both directions follow by induction on $l$, making use of the
  transitivity of $\le$.
\end{proof}

\begin{theorem}[SortedIx\_iff]
  For all lists $l$,
  \begin{equation*}
    \var{Sorted}~l \iff \var{SortedIx}~l
  \end{equation*}
\end{theorem}
\begin{proof}
  We reason by induction on $l$. The base cases are easy, so we
  address the inductive case, $l = a :: l'$

  ($\implies$) We have to prove that $(a :: l')_i \le (a :: l')_j$ given
  $i \le j$ and $\var{Sorted}~l$. We reason by cases on $i$ and $j$:
  \begin{itemize}
  \item $i = j = 0$. By reflexivity, $a \le a$.
  \item $i = 0,~j=j' + 1$. We need to show $a \le l'_{j'}$. We know that
    $a \le x$ for all $x \in l'$ by $\var{Sorted}~(a::l')$, so we are
    done.
  \item $i = i' +1,~j = j' + 1$. We need to show $l'_{i'} \le l'_{j'}$,
    which follows from the induction hypothesis.
  \end{itemize}

@@

The word \textit{permutation} has two related meanings that are often
used interchangably. We sometimes say that two lists are permutations of each
other if they contain all of the same elements, just appearing in a
different order. However, in mathematics we often say that a
permutation is actually the bijection on indices that defines the
correspondence between the two lists. For example, we say that $p$ is
the permutation that relates $l$ to $l'$ if for all indices $i$ and
$j$,
\begin{equation*}
  l_{p(i)} = l'_i
\end{equation*}
For consistency with Coq's built-in \var{Permuation} relation, in a
situation like this we will say that $l$ and $l'$ are permutations of
each other, and that $p$ is the \textit{permutation function} that
relates $l$ to $l'$. We say that two lists $l$ and $l'$ are
permutations of each other is precisely when
\begin{itemize}
  \item $\var{length}~l = \var{length}~l' = n$
  \item there exists a permutation function $p : \mathbb{N}_{<n} \to
    \mathbb{N}_{<n}$, where $p$ is a bijection, such that for all indices $i$,
    \begin{equation*}
      l_{p(i)} = l'_i
    \end{equation*}
\end{itemize}

@@

@@

\begin{proof}
  This follows from \var{StablePerm\_Sorted\_eq}.
\end{proof}

@@

\stablesortunique
\begin{proof}
  This follows from \var{Sort\_StableSort\_Ord} and \var{StableSort\_unique}.
\end{proof}

@@
